---
title: "dplyr examples: happiness"
author: "Heike Hofmann"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

# The Happy data from GSS

The General Social Survey (GSS) has been run by NORC every other year since 1972 to keep track of current opinions across the United States. 

An excerpt of the GSS data called `happy` is available from the `classdata` package:

```{r, eval=FALSE}
remotes::install_github("heike/classdata")
```

```{r}
library(classdata)
head(happy)
```

You can find a codebook with explanations for each of the variables at https://gssdataexplorer.norc.org/

---
class: inverse
# Your Turn

Load the `happy` data from the `classdata` package.

- how many variables, how many observations does the data have? What do the variables mean? 

- Plot the variable `happy`. Introduce a new variable `nhappy` that has values 1 for `not too happy`, 2 for `pretty happy`, 3 for `very happy` and `NA` for missing values. There are multiple ways to get to that. Avoid `for` loops.

- Based on the newly introduced numeric scores, what is the average happiness of respondents?

---
class: inverse
# Your turn

- how does average happiness change over the course of a life time? Is this relationship different for men and women? Draw plots.

- are people now happier than ten years ago? How is happiness related to time?

---
class: inverse
# Your Turn

- Are Republicans or Democrats happier? Compare average happiness levels over `partyid`.

- How are financial relations associated with average happiness levels? Is this association different for men and women?

- Find a plot that shows the differences for each one of the summaries.

---
class: inverse
# Your turn: asking questions 


- What other variable(s) might be associated with happiness? Plot it.


---

# Helper functions (1)

- `n()` provides the number of rows of a subset:
```{r, warning=FALSE, message = FALSE}
library(dplyr)
happy %>% group_by(sex) %>% summarise(n = n())
```

- `tally()` is a combination of `summarise` and `n`
```{r}
happy %>% group_by(sex) %>% tally()
```

---

# Helper functions (2)

- `count()` is a further shortcut of `group_by` and `tally`:

```{r}
happy %>% count(sex, degree)
```

- `count()` doesn't introduce any grouping


---

# Grouping and Ungrouping

- `ungroup` removes a grouping structure from a data set

- necessary to make changes to a grouping variable (such as re-ordering or re-labelling) 


